---
title: '`fields` vignette'
author: "Ashton Wiens, Mitchell Krock, Emma Lilly and Doug Nychka"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
---

```{r set-options, echo=FALSE, eval=TRUE}
options(width = 400)
knitr::opts_chunk$set(fig.show = 'asis')   
# use 'hide' only if you want no figures
library( fields)
```
<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
  color: DarkGreen;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r child="Introduction.Rmd"}
```

\newpage
```{r child="FieldsInAction.Rmd"}
```

\newpage
```{r child="splines.Rmd"}
```

\newpage
```{r child="IntroToSpatialStatistics.Rmd"}
```

\newpage
```{r child="TraditionalGeostatistics.Rmd"}
```

\newpage
```{r child="spatProc.Rmd"}
```
\newpage
```{r child="images.Rmd"}
```

\newpage
```{r child="smooth.Rmd"}
```

\newpage
```{r child="CovarianceRevisited.Rmd"}
```


\newpage
```{r child="mKrig.Rmd"}
```

\newpage
```{r child="Tpsfast.Rmd"}
```

\newpage
```{r child="KrigingTheory.Rmd"}
```
